---
title: R NotebookEDA of NEI records of PM2.5 for 1999, 2002, 2005, and 2008 (part
  of Exploratory Data Analysis by Johns Hopkins University)
output:
  html_document:
    df_print: paged
---


```{r Libraries and loading data, echo=FALSE}
library(tidyr)
library(dplyr)
library(readr)
library(ggplot2)


source("load_data.R")
if (!exists("NEI")){
  data <- load_data()
  NEI <- data[["NEI"]]
  SCC <- data[["SCC"]]
}
```

## Quick look at the data

```{r}
NEI <- as_tibble(NEI)
SCC <- as_tibble(SCC)
str(NEI)
str(SCC)
```

The NEI data is the amount of a specific *Pollutant* for a specific source *SCC* for a *year* in a county indicated by its *FIPS* code.
The *type* represents the type of the emission sources as described [here](https://www.epa.gov/air-emissions-inventories/national-emissions-inventory-nei).


Let's factorize the string values for NEI and check a summary of the NEI.
```{r}
NEI$fips <- parse_factor(NEI$fips)
NEI$SCC <- parse_factor(NEI$SCC, levels = levels(SCC$SCC))
NEI$Pollutant <- parse_factor(NEI$Pollutant)
NEI$type <- parse_factor(NEI$type)
NEI$year <- parse_factor(as.character(NEI$year))
summary(NEI)
```

Similar summary for SCC.

```{r}
summary(SCC)
```

Some values seems missing. Quick overview.

```{r}
summary(is.na(SCC))
```


## Binding the two dataframes

```{r}
data <- NEI %>%
  left_join(SCC, by = c("SCC" = "SCC"))
str(data)
```

## Answering questions with plots

### 1. Have total emissions from PM2.5 decreased in the United States from 1999 to 2008? 

*Using base plotting system*
total PM2.5 emission from all sources for each of the years 1999, 2002, 2005, and 2008
```{r Data preparation for Q1}

totals <- data %>% group_by(year) %>%
  summarise(total = sum(Emissions)/10^6)
```

```{r Data plotting for Q1}
barplot(totals$total, 
        names.arg = totals$year, 
        main = "Emission in millions of tons of PM2.5 from all sources for \n each of the years 1999, 2002, 2005, and 2008",
        xlab = "Year",
        ylab = "Emission of PM2.5 in millions of tons",
        ylim = c(0,8))
abline(lm( totals$total ~as.numeric(totals$year)), col="red")
dev.print(device = png, file = "plot1.png", width = 500, pointsize=10)
```

### 2. Have total emissions from PM2.5 decreased in the Baltimore City, Maryland *(fips == "24510")* from 1999 to 2008? 

*Using base plotting system*

```{r Data preparation for Q2}
totals_baltimore <- data %>% filter(fips == "24510") %>%
  group_by(year) %>%
  summarise(total = sum(Emissions))
```


```{r Data plotting for Q2}
barplot(totals_baltimore$total, 
        names.arg = totals_baltimore$year, 
        main = "Emission in tons of PM2.5 from all sources of Baltimore City \n each of the years 1999, 2002, 2005, and 2008",
        xlab = "Year",
        ylab = "Emission of PM2.5 in tons",
        ylim = (c(0,4000)))

abline(lm( totals_baltimore$total ~as.numeric(totals_baltimore$year)), col="red")
dev.print(device = png, file = "plot2.png", width = 500, pointsize=10)
```


### 3. Of the four types of sources indicated by the *type (point, nonpoint, onroad, nonroad)* variable, which of these four sources have seen decreases in emissions from 1999–2008 for Baltimore City? Which have seen increases in emissions from 1999–2008?


*Using ggplot2 plotting system*

```{r Data preparation for Q3}
totals_baltimore_type <- data %>% filter(fips == "24510") %>%
  group_by(type, year) %>%
  summarise(total = sum(Emissions)) %>%
  mutate(year=as.numeric(levels(year))[year])
```


```{r Data plotting for Q3}
g<-ggplot(totals_baltimore_type) + geom_col(aes(x = year, y = total)) + geom_smooth(method = lm, aes(x = year, y = total), se = FALSE) + facet_grid(rows=vars(type), scales = "free_y")
g <- g + labs(x="Year", y ="Tons of PM2.5", title="Emission for the four types of sources")
g <- g + coord_cartesian(xlim=c(1999,2008))
print(g)
```
```{r echo=FALSE}
ggsave("plot3.png")
```


### 4. Across the United States, how have emissions from coal combustion-related sources changed from 1999–2008?


```{r Data preparation for Q4}
totals_coal <- data %>% filter(EI.Sector == "Fuel Comb - Comm/Institutional - Coal") %>%
  group_by(year) %>%
  summarise(total = sum(Emissions)) %>%
  mutate(year=as.numeric(levels(year))[year])

```


```{r Data plotting for Q4}
g <- ggplot(totals_coal) + geom_col(aes(x = year, y = total)) + geom_smooth(method = lm, aes(x = as.numeric(year), y = total), se = FALSE)
g <- g + labs(x="Year", y ="Tons of PM2.5", title="PM2.5 Emission for coal-combustion (in tons)")
g <- g + coord_cartesian(xlim=c(1999,2008))
print(g)
```
```{r echo=FALSE}
ggsave("plot4.png")
```

### 5. How have emissions from motor vehicle sources changed from 1999–2008 in Baltimore City?

**Hypothesis:** We consider that those emissions are represented by the type *onroad* considering the information for [epa.gov](https://www.epa.gov/air-emissions-inventories/national-emissions-inventory-nei) : "NEI onroad sources include emissions from onroad vehicles that use gasoline, diesel, and other fuels"



```{r Data preparation for Q5}
totals_vehicle <- data %>% filter(type=="ON-ROAD", fips=="24510") %>%
  group_by(year) %>%
  summarise(total = sum(Emissions)) %>%
  mutate(year=as.numeric(levels(year))[year])
```




```{r Data plotting for Q5}
g <- ggplot(totals_vehicle) + geom_col(aes(x = year, y = total)) + geom_smooth(method = lm, aes(x = year, y = total), se = FALSE)
g <- g + labs(x="Year", y ="Tons of PM2.5", title="PM2.5 Emission for motor vehicles sources (in tons)")
g <- g + coord_cartesian(xlim=c(1999,2008),ylim=c(0,400))
print(g)
```

```{r echo=FALSE}
ggsave("plot5.png")
```

### 6. Compare emissions from motor vehicle sources in Baltimore City with emissions from motor vehicle sources in Los Angeles County, California *fips=="06037"*. Which city has seen greater changes over time in motor vehicle emissions?


```{r Data preparation for Q6}
totals_vehicle_comp <- data %>% filter(type=="ON-ROAD", fips=="24510" | fips=="06037") %>%
  group_by(fips,year) %>%
  summarise(total = sum(Emissions)) %>%
  mutate(year=as.numeric(levels(year))[year])
```

```{r echo=FALSE}
# GET EQUATION AND R-SQUARED AS STRING
# SOURCE: https://stackoverflow.com/questions/7549694/add-regression-line-equation-and-r2-on-graph

lm_eqn <- function(df, y, x){
    formula = as.formula(sprintf('%s ~ %s', y, x))
    m <- lm(formula, data=df);
    # formating the values into a summary string to print out
    # ~ give some space, but equal size and comma need to be quoted
    eq <- substitute(italic(target) == a + b %.% italic(input)*","~~italic(r)^2~"="~r2*","~~p~"="~italic(pvalue), 
         list(target = y,
              input = x,
              a = format(as.vector(coef(m)[1]), digits = 2), 
              b = format(as.vector(coef(m)[2]), digits = 2), 
             r2 = format(summary(m)$r.squared, digits = 3),
             # getting the pvalue is painful
             pvalue = format(summary(m)$coefficients[2,'Pr(>|t|)'], digits=5)
            )
          )
    as.character(as.expression(eq));                 
}
```



```{r Data plotting for Q6}
g <- ggplot(totals_vehicle_comp, aes(x = year, y = total, fill=fips))
g <- g + geom_col(position="dodge")
g <- g + geom_smooth(se = FALSE, method = lm)
g <- g + coord_cartesian(xlim=c(1998,2009.5),ylim=c(0,5500))
g <- g + geom_text(aes(x = 1999, y = 5250, hjust="left",label = paste("Baltimore: ", lm_eqn(totals_vehicle_comp %>% filter(fips=="24510"), "total", "year"))), parse = TRUE)
g <- g + geom_text(aes(x = 1999, y = 5000, hjust="left", label = paste("LA: ", lm_eqn(totals_vehicle_comp %>% filter(fips=="06037"), "total", "year"))), parse = TRUE)
g <- g + scale_fill_discrete(name="County", labels=c("Baltimore City", "Los Angeles County"), aesthetics = c("colour", "fill"))
g <- g + labs(x="Year", y ="Tons of PM2.5", title="PM2.5 Emission for motor vehicles sources in Baltimore City and Los Angeles County (in tons)")
print(g)
```
```{r echo=FALSE}
ggsave("plot6.png")
```















