---
title: "Developing Data Products Project - France Birth Data"
author: "Baptiste Maingret"
date: "20/11/2019"
output: 
  ioslides_presentation: 
    smaller: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Purpose and functionning

### Purpose
The goal of this application is to see the evolution of the France birth rate and be able to quickly visualize high birth rate region on a map for a particular year.

### Functionning

The user can select the year to select the data to be displayed on the map.

At any time the user can see the overall plot of the data on the chart below the slider.


## Data set

All data preprocessing can be found in the *data_process.R* file.

### Birth data

Data comes from the French national agency for statistics [INSEE](https://www.insee.fr/fr/statistiques/series/102928992).

It came as multiple files, so some preprocessing was required.

### Regions data

To be able to draw the regions on maps a file describing the contour of each regions was required.
The region polygons can be found on [GitHub](https://raw.githubusercontent.com/gregoiredavid/france-geojson).



## Plotting geographical regions on map

The plotting of the regions was a good challenge.

It requires a specific file format (geojson), a specific library(geojsonio) to generate an object of the *Spatial class*:

```{r echo=TRUE, eval=FALSE}
  regions <- geojson_read("https://raw.githubusercontent.com/gregoiredavid/france-geojson/master/regions-version-simplifiee.geojson", what="sp")
```

The plotting thanks to leaflet was also challenging, especially to be able to dynamically change the year of the data to be displayed.
```{r echo=TRUE, eval=FALSE}
leaflet(regions) %>%
            addTiles() %>%
            addPolygons(
                stroke = FALSE,
                smoothFactor = 0.3,
                fillOpacity = 1,
                fillColor = as.formula(paste0("~pal(y", input$year, ")")),
                label = as.formula(paste0("~paste0(y", input$year, ")"))
            ) 
```


## Improvements

Some possible future improvements

  * Animated maps that scroll automatically through the year
  * Regression model fitted to the data to have quantitive information on evolution
  * Add death data and be able to switch between the two
  