---
title: "Activity monitoring data analysis"
output: 
  html_document:
    keep_md: yes
  github_document: default
---

```{r library, echo=FALSE,results='hide',warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
```

## Activity monitoring (part of Reproducible Research module by Johns Hopkins University)

This assignment was part of the Johns Hopkins Coursera module on [Reproducible Research](https://www.coursera.org/learn/reproducible-research) as part of the [Data Sciene Specialization](https://www.coursera.org/specializations/jhu-data-science).


Full code can be found on [GitHub](https://github.com/bmaingret/RepData_PeerAssessment1).

## Loading and preprocessing the data

The variables included in this dataset are:

* **steps**: Number of steps taking in a 5-minute interval (missing
    values are coded as `NA`)

* **date**: The date on which the measurement was taken in YYYY-MM-DD
    format

* **interval**: Identifier for the 5-minute interval in which
    measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there
are a total of 17,568 observations in this
dataset.

Loading the data:
```{r}
if (!file.exists("activity.csv")){
  unzip("activity.zip")
}
data <- read.csv("activity.csv", na.strings = "NA", colClasses = c("integer", "character", "integer"))
data$date <- as.Date(data$date, format="%Y-%m-%d")
```

Checking the data:
```{r}
str(data)
summary(data)
head(data)
```

## What is mean total number of steps taken per day?
*For this part of the assignment, you can ignore the missing values in the dataset.*

Summarizing the data:
```{r}
total_steps <- data %>% group_by(date) %>%
  summarise(total = sum(steps, na.rm = TRUE))
```
```{r}
g <- ggplot(total_steps, aes(total))
g <- g + geom_histogram()
g <- g + labs(title="Number of steps per day", y="", x="")
g
```

```{r}
mean <- mean(total_steps$total, na.rm = TRUE)
mean
```
The mean total number of steps per day is: **`r format(mean, nsmall=2)`**

```{r}
median <- median(total_steps$total, na.rm = TRUE)
median
```
The median total number of steps per day is: **`r median`**

## What is the average daily activity pattern?

Summarizing the data:
```{r}
steps_interval <- data %>% group_by(interval) %>%
  summarise(mean = mean(steps, na.rm = TRUE))
```

```{r}

g <- ggplot(steps_interval, aes(interval, mean))
g <- g + geom_line()
g <- g + labs(title="Average steps per interval", y="Average steps", x="Interval")
g
```

```{r}
ix <- which.max(steps_interval$mean)
interval <- steps_interval[[ix, "interval"]]
val <- steps_interval[[ix, "mean"]]
```

The interval with max mean number of steps is **`r interval`** with a mean number of steps of **`r format(val, digit=2, nsmall = 2)`**.
```{r}
h <- floor(interval/60)
m <- interval%%60
```

Supposing the interval starts at 00:00 of each day, this interval corresponds to **`r paste(h,m, sep=":")`**.

## Imputing missing values

Total number of missing values:
```{r}
apply(is.na(data), 2, sum)
```

We will fill in the missing steps values with the mean for the specific day and interval.

First we compute the mean for each interval and day of the week.
```{r}
data <- data %>% mutate(weekday = as.factor(as.POSIXlt(date)$wday))
fill_val <- data %>% group_by(weekday, interval) %>%
  summarise(mean = mean(steps, na.rm = TRUE))
```

Imputing missing data.
```{r}
data_nna <- data

for (row in 1:nrow(data_nna)) {
  if(is.na(data_nna[row, "steps"])) {
    wd <- data_nna[row, "weekday"]
    interval <- data[row, "interval"]
    data_nna[row, "steps"] <- fill_val[fill_val$weekday==wd & fill_val$interval==interval, "mean"]
  }
}

apply(is.na(data_nna), 2, sum)

```

Repeating first steps of the assignement now with the imputed data.
Summarizing the data:
```{r}
total_steps_nna <- data_nna %>% group_by(date) %>%
  summarise(total = sum(steps))
```
```{r}
g <- ggplot(total_steps_nna, aes(total))
g <- g + geom_histogram()
g <- g + labs(title="Number of steps per day", y="", x="")
g
```

```{r}
mean_nna <- mean(total_steps_nna$total)
mean_nna
```
The mean total number of steps per day is: **`r format(mean_nna, nsmall=2)`** (was `r format(mean, nsmall=2)` before imputation).

```{r}
median_nna <- median(total_steps_nna$total)
median_nna
```
The median total number of steps per day is: **`r format(median_nna, nsmall=2)`** (was `r format(median, nsmall=2)` before imputation).

## Are there differences in activity patterns between weekdays and weekends?

Summarizing the data:
```{r}
steps_interval_nna <- data_nna %>% mutate(week.part = if_else(weekday %in% c(1,6), "weekend", "weekdays")) %>%
  group_by(week.part,interval) %>%
  summarise(mean = mean(steps, na.rm = TRUE))
```


```{r}
g <- ggplot(steps_interval_nna, aes(interval, mean, col=week.part))
g <- g + geom_line()
g <- g + facet_grid(rows = vars(week.part))
g <- g + labs(title="Average steps per interval between weekdays and week", y="Average steps", x="Interval")
g
```

