---
title: "EDA of USA National Weather Service Storm Data"
author: "Baptiste Maingret"
date: "23/10/2019"
output:
  
  github_document: default
  html_document: 
    keep_md: no
---

## EDA of U.S. National Oceanic and Atmospheric Administration's storm database (part of Reproducible Research Analysis by Johns Hopkins University)

This assignment was part of the Johns Hopkins Coursera module on [Reproducible Research](https://www.coursera.org/learn/reproducible-research) as part of the [Data Sciene Specialization](https://www.coursera.org/specializations/jhu-data-science).

 <!--more-->

Source code available on [GitHub](https://github.com/bmaingret/eda-storm-data)

[National Weather Service Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

[National Climatic Data Center Storm Events FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)


## Synopsis 

The purpose of this analysis is to answer two questions:  

  1. Across the United States, which types of events are most harmful with respect to population health?
  2. Across the United States, which types of events have the greatest economic consequences?

For the first question, we need to quantify *harmful* considering the data we have at hands. Two variables can be seen as good candidates:  

  * injuries
  * casualties


To simplify our analysis we will consider that both variables account for the same degree of harm to population health, and that the count of injuries and casualties is a good indicator.



For the second question, we need to quantify *economic consequences* considering the data we have at hands. Two variables can be seen as good candidates (associated with their explanoratory variables):  

  * propdmg
  * cropdmg

These variables summarize estimates of damage in dollars amount, and we will consider the sum of these to be a good indicator for our purpose.


## Data Processing
```{r Libraries, message=FALSE}
library(dplyr)
library(tidyr)
library(forcats)
library(ggplot2)
library(knitr)
```

### Loading data
```{r Data loading, cache=TRUE, warning=FALSE, message=FALSE}
URL = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
FILE = "stormdata.csv.bz2"
if (!file.exists(FILE)) {
  download.file(URL, FILE, method="libcurl")
}

storm.data <- read.csv(FILE)
colnames(storm.data) <- make.names(casefold(colnames(storm.data), upper = FALSE), unique=TRUE, allow_ = FALSE)
```

### Glimpse at data
```{r Data summary, cache=TRUE}
head(storm.data)
str(storm.data)
```

### 1. Across the United States, which types of events are most harmful with respect to population health?

For first question we will use the *injuries* and *fatalities* columns which do not appear to have any specific or impacting issues.
```{r "Injuries and Fatalities"}
summary(storm.data[,c("injuries", "fatalities")])
sum(is.na(storm.data[,c("injuries", "fatalities")]))
```

We take the assumption that the number of injuries and fatalities characterizes the harm to population health in equal weight.
Thus we look for the events for which the cumulated number of injuries and fatalities are the highest.


Let's summarize data accordingly:
```{r "Data prep for Q1"}
data.1 <- storm.data %>% group_by(evtype) %>%
  summarise_at(vars(fatalities, injuries), sum) %>%
  top_n(5, fatalities + injuries) %>%
  mutate(evtype = fct_drop(evtype)) %>%
  mutate(evtype = fct_reorder(evtype, fatalities + injuries, .desc = TRUE)) %>%
  pivot_longer(c(fatalities, injuries), names_to="casualty.type", values_to ="casualties")
```

### 2. Across the United States, which types of events have the greatest economic consequences?

For the second question we will use the *propdmg*, *prodmgexp*, *cropdmg* and *cropdmgexp* columns (we could drop the crop variables considering the values).

```{r "Expenses"}
summary(storm.data$propdmgexp)

non.char.levels <- levels(storm.data$propdmgexp)[!grepl("[A-z]", levels(storm.data$propdmgexp))]
char.levels <- levels(storm.data$propdmgexp)[grepl("[A-z]", levels(storm.data$propdmgexp))]

storm.data %>% mutate(dmg.level = propdmgexp %in% non.char.levels) %>%
  group_by(dmg.level) %>%
  summarise(total = sum(propdmg))

# Similar work for the cropdmg exp
storm.data %>% mutate(dmg.level = cropdmgexp %in% non.char.levels) %>%
  group_by(dmg.level) %>%
  summarise(total = sum(cropdmg))
```
There is not much information on the  *prodmgexp* and *cropdmgexp* variables.  
When we look at those variables, we can find what looks like standard M/K/B SI units (in addition to H).
We will chose to ignore levels that are not one of *`r char.levels`*, considering the lack of information and the fact that the associated amount are limited compared to the rest.


Let's summarize data accordingly:
```{r "Data prep for Q2"}
data.2 <- storm.data %>% group_by(evtype) %>%
  filter(propdmgexp %in% char.levels | cropdmgexp %in% char.levels) %>%
  mutate(propdmgexp = casefold(propdmgexp, upper=TRUE), cropdmgexp = casefold(cropdmgexp, upper=TRUE)) %>%
  mutate(propdmg = case_when(
    propdmgexp == "H" ~ propdmg*100,
    propdmgexp == "K" ~ propdmg*1000,
    propdmgexp == "M" ~ propdmg*1000000,
    propdmgexp == "B" ~ propdmg*1000000000,
  )) %>%
  mutate(cropdmg = case_when(
    cropdmgexp == "H" ~ cropdmg*100,
    cropdmgexp == "K" ~ cropdmg*1000,
    cropdmgexp == "M" ~ cropdmg*1000000,
    cropdmgexp == "B" ~ cropdmg*1000000000,
  )) %>%  
  summarise_at(vars(propdmg, cropdmg), sum) %>%
  top_n(5, propdmg + cropdmg) %>%
  mutate(evtype = fct_drop(evtype)) %>%
  mutate(evtype = fct_reorder(evtype, propdmg + cropdmg, .desc = TRUE)) %>%
  pivot_longer(c(propdmg, cropdmg), names_to="damage.type", values_to ="amount")
```

## Results

### 1. Across the United States, which types of events are most harmful with respect to population health?

we plot the top 5  type of events considering the number of casualties.  
We can see that the tornadoes are #1 by order of magnitudes.


```{r "Plot for Q1"}
g.1 <- ggplot(data.1, aes(evtype, casualties, fill=casualty.type))
g.1 <- g.1 + geom_col(position="stack")
g.1 <- g.1 + labs(title="Top 5 natural events for number of casualties \n (combine injuries and fatalities)", x="Event type", y="Number of casualties")
g.1 <- g.1 + scale_fill_discrete(name="Type of casualty")
g.1 <- g.1 + coord_flip()
g.1
```


### 2. Across the United States, which types of events have the greatest economic consequences?

we plot the top 5  type of events considering the amount of damages.  
We can see again that the tornadoes are #1 by order of magnitudes.

```{r "Plotfor Q2"}
g.2 <- ggplot(data.2, aes(evtype, amount/1000000, fill=damage.type))
g.2 <- g.2 + geom_col(position="stack")
g.2 <- g.2 + labs(title="Top 5 natural events for damage estimates \n (combine crop and properties)", x="Event type", y="Damage estimate (millions of dollars)")
g.2 <- g.2 + scale_fill_discrete(name="Type of damage")
g.2 <- g.2 + coord_flip()
g.2
```


