---
title: "Effect of Vitamin C on Tooth Growth in Guinea Pigs"
author: "Baptiste Maingret"
category: [datascience, assignement]
tags: [R, EDA, Data Science, Coursera, Plot, Statistics]
output: 
  github_document: default
  pdf_document: default
---



## Basic Inferential Data Analysis on ToothGrowth dataset (part of Statistical Inference by Johns Hopkins University)

This assignment was part of the Johns Hopkins Coursera module on
[Statistical Inference](https://www.coursera.org/learn/statistical-inference) as part
of the [Data Science
Specialization](https://www.coursera.org/specializations/jhu-data-science).

<!--more-->

Source code available on [GitHub](https://github.com/bmaingret/stat-inference-project)

## Overview

The goal is to conduct some simple hypothesis testing on the ToothGrowth dataset available in the R datasets package.

Some assumptions:

  * equal variances among groups
  * standard deviation estimated from the samples
  * $\alpha$ is set to 5%
  * samples are not paired

## Data processing

We import the data and directly set the *dose* as a factor.
```{r}
library(ggplot2)
library(datasets)
tg <- datasets::ToothGrowth
tg$dose <- as.factor(tg$dose)
```


Glimpse at data.
```{r}
str(tg)
summary(tg)
```

Some plots.

```{r}
  qplot(x=len, data=tg, color=dose, group = dose, geom = "density", facets = dose ~ supp)
```

### Has the delivery method an impact on tooth growth?

We will test in regards of the null-hypothesis that their is no difference in means between the two groups.

```{r}
n = 10
x = tg[tg$supp=="OJ", "len"]
y = tg[tg$supp=="VC", "len"]
delta = mean(x) - mean(y)
p.sd = sqrt((var(x)+var(y))/2)

t.res <- t.test(x, y, alternative = "two.sided", mu = 0, paired = FALSE, var.equal = TRUE)
p.res <- power.t.test(n, delta, p.sd, sig.level=0.05, type="two.sample", alternative="two.sided")
```

We have a p-value (`r t.res$p.value*100`%) larger the 5% and in addition the confidence interval (`r t.res$conf.int`) covers the value 0. We fail to reject the null hypothesis in this case.

### Has the dose an impact on tooth growth?

We test the difference in means between each dosage (3 tests: 0.05 vs 1, 0.5 vs 2, 1 vs 2).

```{r}
n = 10
x = tg[tg$dose=="0.5", "len"]
y = tg[tg$dose=="1", "len"]
delta = mean(x) - mean(y)
p.sd = sqrt((var(x)+var(y))/2)

t.res.a <- t.test(x, y, alternative = "two.sided", mu = 0, paired = FALSE, var.equal = TRUE)
p.res.a <- power.t.test(n, delta, p.sd, sig.level=0.05, type="two.sample", alternative="two.sided")
```


```{r}
n = 10
x = tg[tg$dose=="0.5", "len"]
y = tg[tg$dose=="2", "len"]
delta = mean(x) - mean(y)
p.sd = sqrt((var(x)+var(y))/2)

t.res.b <- t.test(x, y, alternative = "two.sided", mu = 0, paired = FALSE, var.equal = TRUE)
p.res.b <- power.t.test(n, delta, p.sd, sig.level=0.05, type="two.sample", alternative="two.sided")
```


```{r}
n = 10
x = tg[tg$dose=="1", "len"]
y = tg[tg$dose=="2", "len"]
delta = mean(x) - mean(y)
p.sd = sqrt((var(x)+var(y))/2)

t.res.c <- t.test(x, y, alternative = "two.sided", mu = 0, paired = FALSE, var.equal = TRUE)
p.res.c <- power.t.test(n, delta, p.sd, sig.level=0.05, type="two.sample", alternative="two.sided")
```
```{r echo=FALSE}
a <- c(t.res.a$p.value, t.res.a$conf.int, p.res.a$power)
b <- c(t.res.b$p.value, t.res.b$conf.int, p.res.b$power)
c <- c(t.res.c$p.value, t.res.c$conf.int, p.res.c$power)
  
results <- data.frame(dose.0.5v1=a,dose.0.5v2=b, dose.1v2=c,row.names = c("p-value", "conf-interval-low", "conf-interval-up", "power"))
results["p-value",] <- results["p-value",]
results
```


## Conclusions

We failed to reject the null-hypothesis regarding the impact of the delivery method on tooth growth.

The dosage was found to be statistically significant and tests rejected the null-hypothesis.